---
title: "UK elevation and seasonal temperature"
output:
  html_document:
    theme: null
    mathjax: null
jupyter:
  jupytext:
    formats: ipynb,Rmd
    main_language: R
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.19.2
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyterra)
library(terra)
library(geodata)
library(rnaturalearth)
library(ggplot2)
library(dplyr)
library(sf)
library(rlang)
```

## Introduction

A **SpatRaster** from the **terra** package stores one or more gridded layers. Each layer is a matrix of cell values aligned to a common extent, resolution, and **coordinate reference system (CRS)**. When a raster has multiple layers—such as twelve monthly temperature grids—each layer is like a variable measured at the same locations on the Earth's surface.

**tidyterra** extends **dplyr**-style verbs to **SpatRaster** and **SpatVector** objects. Unlike applying **dplyr** to a plain data frame, **`tidyterra::mutate()`**, **`select()`**, and **`rename()`** on a raster return another **SpatRaster** (or updated raster), preserving spatial metadata. That keeps workflows inside the **terra** object model instead of melting everything to long tables first.

```{r load-elevation, message=FALSE}
uk_elev <- geodata::elevation_30s(country = "GBR", path = tempdir())
uk_boundary <- rnaturalearth::ne_countries(
  country = "United Kingdom",
  scale = "medium",
  returnclass = "sv"
)

if (!terra::same.crs(uk_elev, uk_boundary)) {
  uk_boundary <- terra::project(uk_boundary, uk_elev)
}

uk_elev_cropped <- terra::crop(uk_elev, uk_boundary)
uk_elev_masked <- terra::mask(uk_elev_cropped, uk_boundary)

elev_layer <- names(uk_elev_masked)[1]
uk_elev_masked <- uk_elev_masked |>
  tidyterra::rename(elevation = !!rlang::sym(elev_layer)) |>
  tidyterra::mutate(
    elev_band = cut(
      elevation,
      breaks = c(-Inf, 100, 300, 600, Inf),
      labels = c("0-100 m", "100-300 m", "300-600 m", "600+ m"),
      include.lowest = TRUE
    )
  )
```

The code above downloads a **30 arc-second** (~1 km) elevation grid for Great Britain, clips it to the Natural Earth **United Kingdom** polygon, and uses **`terra::crop`** and **`terra::mask`** so only cells inside the national boundary contribute to the map. (Geometric clipping lives in **terra**; **tidyterra** is used for **dplyr**-style layer operations such as **`rename`** and **`mutate`**.) A **numeric** layer `elevation` (metres) is used for continuous mapping; **`elev_band`** is a derived categorical layer created with **`cut()`** inside **`mutate()`**, illustrating dplyr-like syntax on a **SpatRaster** without dropping the grid structure.

**Hypsometric tint** palettes mimic cartographic conventions for relief: low elevations read as greens and yellows, high ground as browns and whites. That perceptual ordering matches how people interpret terrain on topographic maps, so **`scale_fill_hypso_tint_c()`** is a natural choice for continuous elevation—even when we also keep discrete **bands** for other summaries.

```{r plot-elevation, message=FALSE, fig.cap="UK elevation (masked) with country outline."}
p_elev <- ggplot() +
  tidyterra::geom_spatraster(data = uk_elev_masked, aes(fill = elevation)) +
  tidyterra::scale_fill_hypso_tint_c(
    name = "m a.s.l.",
    guide = guide_colorbar(barwidth = unit(3.5, "cm"), title.position = "top")
  ) +
  tidyterra::geom_spatvector(data = uk_boundary, fill = NA, colour = "grey20", linewidth = 0.35) +
  labs(title = "UK Elevation") +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", margin = margin(b = 6)),
    legend.title = element_text(size = 9),
    legend.position = "bottom"
  )

p_elev
```

## Seasonal mean temperature

WorldClim monthly mean temperature for the UK arrives as a **twelve-layer SpatRaster** (one layer per month). Layers are indexed **1–12** for January–December. We keep four representative months—**December (12), March (3), June (6), September (9)**—as proxies for meteorological **winter, spring, summer, and autumn** in the Northern Hemisphere.

```{r worldclim, message=FALSE}
tavg <- geodata::worldclim_country(country = "GBR", var = "tavg", res = 10, path = tempdir())

layer_names <- names(tavg)
season_pick <- layer_names[c(12L, 3L, 6L, 9L)]

uk_tavg_seasons <- tavg |>
  tidyterra::select(dplyr::all_of(season_pick)) |>
  tidyterra::rename(
    Winter = !!rlang::sym(season_pick[1]),
    Spring = !!rlang::sym(season_pick[2]),
    Summer = !!rlang::sym(season_pick[3]),
    Autumn = !!rlang::sym(season_pick[4])
  )
```

Here **`tidyterra::select()`** subsets raster **layers** by name (analogous to choosing columns). **`rename()`** gives human-readable season labels while the result remains a **SpatRaster** suitable for **`geom_spatraster()`** and faceting.

The **whitebox**-derived continuous palettes (here **`"muted"`**) provide colour ramps tuned for terrain and climate-style surfaces; they offer an alternative to hypsometric tints when the variable is **temperature** rather than relief.

```{r plot-temperature, message=FALSE, fig.cap="Mean temperature (°C) by season (Dec, Mar, Jun, Sep)."}
ggplot() +
  tidyterra::geom_spatraster(data = uk_tavg_seasons, aes(fill = after_stat(value))) +
  facet_wrap(~lyr, ncol = 2) +
  tidyterra::scale_fill_whitebox_c(
    palette = "muted",
    name = "°C",
    guide = guide_colorbar(barwidth = unit(2.8, "cm"), title.position = "top")
  ) +
  tidyterra::geom_spatvector(data = uk_boundary, fill = NA, colour = "grey35", linewidth = 0.25) +
  labs(title = "UK Mean Temperature by Season (°C)") +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", margin = margin(b = 8)),
    strip.text = element_text(face = "bold", size = 10),
    legend.title = element_text(size = 9),
    legend.position = "bottom"
  )
```

Together, the elevation and temperature maps show how **multi-layer rasters** encode **space × variable × time** (or season), and how **tidyterra** keeps those structures intact while integrating with **ggplot2**.
