---
title: "English Premiership grounds on UK terrain"
output:
  html_document:
    theme: null
    mathjax: null
jupyter:
  jupytext:
    formats: ipynb,Rmd
    main_language: R
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.19.2
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyterra)
library(terra)
library(geodata)
library(rnaturalearth)
library(ggplot2)
library(dplyr)
library(sf)
library(rlang)
library(knitr)
```

## Reloading the elevation surface

This notebook **rebuilds** the UK elevation raster the same way as in `01_terrain_analysis.Rmd`, so it can be rendered on its own. In a larger project you might save an intermediate **GeoTIFF** under `data/` and read it with **`terra::rast()`** to avoid duplicate downloads.

```{r elevation-rebuild, message=FALSE}
uk_elev <- geodata::elevation_30s(country = "GBR", path = tempdir())
uk_boundary <- rnaturalearth::ne_countries(
  country = "United Kingdom",
  scale = "medium",
  returnclass = "sv"
)

if (!terra::same.crs(uk_elev, uk_boundary)) {
  uk_boundary <- terra::project(uk_boundary, uk_elev)
}

uk_elev_masked <- terra::mask(terra::crop(uk_elev, uk_boundary), uk_boundary)
elev_layer <- names(uk_elev_masked)[1]
uk_elev_masked <- uk_elev_masked |>
  tidyterra::rename(elevation = !!rlang::sym(elev_layer))
```

## Club locations and CRS

Premiership grounds are stored as a **data frame** of **longitude** and **latitude** in **decimal degrees**. **`EPSG:4326`** is the standard **WGS84** geographic CRS used by GPS and most global rasters (including the elevation and WorldClim grids used here): positions are angles on the ellipsoid, not metres on a flat map.

```{r clubs}
rclubs <- data.frame(
  club = c(
    "Bath Rugby", "Bristol Bears", "Exeter Chiefs", "Gloucester Rugby",
    "Harlequins", "Leicester Tigers", "Newcastle Falcons",
    "Northampton Saints", "Sale Sharks", "Saracens"
  ),
  lon = c(
    -2.361, -2.571, -3.525, -2.243, -0.215, -1.133,
    -1.622, -0.906, -2.338, -0.100
  ),
  lat = c(
    51.381, 51.455, 50.730, 51.866, 51.481, 52.620,
    54.961, 52.241, 53.419, 51.674
  ),
  region = c(
    "South West", "South West", "South West", "South West",
    "London", "Midlands", "North",
    "Midlands", "North", "London"
  )
)

clubs_sv <- terra::vect(rclubs, geom = c("lon", "lat"), crs = "EPSG:4326")
```

## Raster–point extraction

**`terra::extract()`** performs a **spatial intersection** between a raster and a set of **sample locations** (points, polygons, or another raster). For points, each location receives the **cell value(s)** of the raster layer(s) at the containing cell (by default the cell **centre** rule applies for points). That is how we attach **terrain height** at each stadium to the club table.

```{r extract}
elev_at_clubs <- terra::extract(uk_elev_masked, clubs_sv)

clubs_elev <- rclubs |>
  mutate(ground_elevation_m = elev_at_clubs[["elevation"]])

clubs_elev |>
  arrange(desc(ground_elevation_m)) |>
  kable(digits = 1, caption = "Premiership grounds ranked by extracted elevation (m)")
```

Geographically, **south-western** clubs such as **Exeter Chiefs** sit on higher ground than **London** sides (**Harlequins**, **Saracens**), which lie in the low-lying **Thames** basin. **Newcastle Falcons** in the north-east can be relatively high or low depending on fine-scale coastal versus inland cells, but the table makes the **ranking** explicit from the grid we used.

The bar chart below uses the same extracted elevations, with clubs ordered from **lowest** to **highest** ground so you can read off the full spread at a glance (still tied to the SRTM-scale raster cells used above).

```{r bar-elevation, message=FALSE, fig.cap="Extracted stadium elevation (m a.s.l.) by club, lowest to highest."}
clubs_elev_bar <- clubs_elev |>
  mutate(club = stats::reorder(club, ground_elevation_m))

ggplot(clubs_elev_bar, aes(x = club, y = ground_elevation_m, fill = region)) +
  geom_col(width = 0.78, colour = "grey35", linewidth = 0.25) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2", name = "Region") +
  labs(
    title = "Premiership ground elevation (ascending)",
    x = NULL,
    y = "Elevation (m a.s.l.)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0, margin = margin(b = 8)),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

```{r map-overlay, message=FALSE, fig.cap="Premiership grounds overlaid on UK elevation."}
ggplot() +
  tidyterra::geom_spatraster(data = uk_elev_masked, aes(fill = elevation)) +
  tidyterra::scale_fill_hypso_tint_c(
    name = "m a.s.l.",
    guide = guide_colorbar(barwidth = unit(3.5, "cm"), title.position = "top")
  ) +
  tidyterra::geom_spatvector(data = clubs_sv, aes(colour = region), size = 3, show.legend = TRUE) +
  tidyterra::geom_spatvector_label(
    data = clubs_sv,
    aes(label = club),
    size = 2.4,
    colour = "grey10",
    label.size = 0.15,
    show.legend = FALSE
  ) +
  tidyterra::geom_spatvector(data = uk_boundary, fill = NA, colour = "grey15", linewidth = 0.3) +
  scale_colour_brewer(palette = "Set2", name = "Region") +
  labs(title = "English Premiership Rugby Grounds by Elevation") +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", margin = margin(b = 6)),
    legend.title = element_text(size = 9),
    legend.position = "bottom"
  )
```

**`geom_spatvector_label()`** draws text from **SpatVector** attributes; where labels crowd together in static output, **ggrepel** could be swapped in with an **sf** conversion for automatic jittering.
