---
title: "Super Rugby — stadium elevation"
output:
  html_document:
    theme: null
    mathjax: null
jupyter:
  jupytext:
    formats: ipynb,Rmd
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.2'
      jupytext_version: 1.19.2
  kernelspec:
    display_name: R
    language: R
    name: ir
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyterra)
library(terra)
library(geodata)
library(rnaturalearth)
library(ggplot2)
library(dplyr)
library(sf)
library(rlang)
library(knitr)
library(patchwork)
library(leaflet)
library(htmlwidgets)
library(htmltools)
library(raster)
library(IRdisplay)
```

## Super Rugby (New Zealand, Australia, South Africa)

**Super Rugby** has taken several formats. This notebook focuses on the classic **three-country** footprint—franchises in **New Zealand**, **Australia**, and **South Africa** that shared the competition before South African sides moved to the **United Rugby Championship**. **Super Rugby Pacific** today is centred on New Zealand and Australia; the South African grounds here are the same venues those franchises used in the Super Rugby era.

Elevation comes from **`geodata::elevation_30s()`** tiles for **`NZL`**, **`AUS`**, and **`ZAF`**, with each stadium sampled from its national raster (same pattern as the URC notebook). **Coordinates** are approximate stadium centre points; edit `super_clubs` to refine them.

```{r super-table}
super_clubs <- data.frame(
  club = c(
    "Blues", "Chiefs", "Crusaders", "Highlanders", "Hurricanes",
    "Brumbies", "Reds", "Waratahs", "Western Force", "Rebels",
    "Bulls", "Stormers", "Sharks", "Lions"
  ),
  stadium = c(
    "Eden Park", "FMG Stadium Waikato", "Apollo Projects Stadium",
    "Forsyth Barr Stadium", "Sky Stadium",
    "GIO Stadium", "Suncorp Stadium", "Allianz Stadium",
    "HBF Park", "AAMI Park",
    "Loftus Versfeld", "DHL Stadium", "Hollywoodbets Kings Park",
    "Emirates Airline Park"
  ),
  lon = c(
    174.744, 175.272, 172.636, 170.526, 174.776,
    149.002, 153.009, 151.225, 115.770, 144.983,
    28.229, 18.411, 31.030, 28.061
  ),
  lat = c(
    -36.875, -37.781, -43.531, -45.869, -41.272,
    -35.250, -27.464, -33.889, -31.938, -37.825,
    -25.748, -33.903, -29.825, -26.198
  ),
  country = c(
    rep("New Zealand", 5),
    rep("Australia", 5),
    rep("South Africa", 4)
  ),
  elev_source = c(
    rep("NZL", 5),
    rep("AUS", 5),
    rep("ZAF", 4)
  ),
  stringsAsFactors = FALSE
)

super_sv <- terra::vect(super_clubs, geom = c("lon", "lat"), crs = "EPSG:4326")
```

## Download elevation tiles

```{r elev-tiles, message=FALSE}
td <- tempdir()
elev_tiles <- list(
  NZL = geodata::elevation_30s(country = "NZL", path = td),
  AUS = geodata::elevation_30s(country = "AUS", path = td),
  ZAF = geodata::elevation_30s(country = "ZAF", path = td)
)

for (nm in names(elev_tiles)) {
  names(elev_tiles[[nm]]) <- paste0("elev_", nm)
}
```

## Extract elevation per ground

```{r extract-super}
extract_point_elev <- function(lon, lat, src) {
  r <- elev_tiles[[src]]
  pts <- terra::vect(
    data.frame(lon = lon, lat = lat),
    geom = c("lon", "lat"),
    crs = "EPSG:4326"
  )
  if (!terra::same.crs(r, pts)) {
    pts <- terra::project(pts, r)
  }
  ex <- terra::extract(r, pts)
  as.numeric(ex[[names(r)[1]]])
}

super_elev <- super_clubs |>
  mutate(
    ground_elevation_m = mapply(
      extract_point_elev,
      lon,
      lat,
      elev_source,
      USE.NAMES = FALSE
    )
  )

super_elev |>
  arrange(desc(ground_elevation_m)) |>
  knitr::kable(
    digits = 1,
    caption = "Super Rugby grounds ranked by extracted elevation (m a.s.l.; national SRTM tile)"
  )
```

```{r bar-super, message=FALSE, fig.cap="Super Rugby stadium elevation (m a.s.l.) by franchise, ascending."}
super_bar <- super_elev |>
  mutate(club = stats::reorder(club, ground_elevation_m))

ggplot(super_bar, aes(x = club, y = ground_elevation_m, fill = country)) +
  geom_col(width = 0.78, colour = "grey35", linewidth = 0.25) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2", name = "Country") +
  labs(
    title = "Super Rugby ground elevation (ascending)",
    x = NULL,
    y = "Elevation (m a.s.l.)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0, margin = margin(b = 8)),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

**Bulls** and **Lions** again sit on the **Highveld** and pull the vertical scale upward, masking smaller differences among New Zealand and Australian venues (often coastal or low-lying basins). The chart below drops those two franchises so the **remaining twelve** grounds are easier to compare.

```{r bar-super-no-highveld, message=FALSE, fig.cap="Super Rugby elevation (ascending) excluding Bulls and Lions."}
super_bar_trim <- super_elev |>
  dplyr::filter(!club %in% c("Bulls", "Lions")) |>
  mutate(club = stats::reorder(club, ground_elevation_m))

ggplot(super_bar_trim, aes(x = club, y = ground_elevation_m, fill = country)) +
  geom_col(width = 0.78, colour = "grey35", linewidth = 0.25) +
  coord_flip() +
  scale_fill_brewer(palette = "Set2", name = "Country") +
  labs(
    title = "Super Rugby elevation (ascending), excluding Bulls & Lions",
    x = NULL,
    y = "Elevation (m a.s.l.)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0, margin = margin(b = 8)),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

## Regional maps (New Zealand, Australia, South Africa)

We **crop** each national DEM to the stadium footprint, **merge** New Zealand and Australia for one panel, and map **South Africa** separately.

```{r map-prep, message=FALSE}
world_sf <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")
super_outline_sf <- world_sf[world_sf$iso_a3 %in% c("NZL", "AUS", "ZAF"), ]
super_outline <- terra::vect(super_outline_sf)

anz_df <- super_clubs |> dplyr::filter(elev_source %in% c("NZL", "AUS"))
sa_df <- super_clubs |> dplyr::filter(elev_source == "ZAF")

anz_ext <- terra::extend(
  terra::ext(terra::vect(anz_df, geom = c("lon", "lat"), crs = "EPSG:4326")),
  2
)
sa_ext <- terra::extend(
  terra::ext(terra::vect(sa_df, geom = c("lon", "lat"), crs = "EPSG:4326")),
  2
)

r_anz <- terra::merge(
  terra::crop(elev_tiles$NZL, anz_ext),
  terra::crop(elev_tiles$AUS, anz_ext)
)
names(r_anz) <- "elevation"

r_sa <- terra::crop(elev_tiles$ZAF, sa_ext)
names(r_sa) <- "elevation"

anz_pts <- terra::vect(anz_df, geom = c("lon", "lat"), crs = "EPSG:4326")
sa_pts <- terra::vect(sa_df, geom = c("lon", "lat"), crs = "EPSG:4326")

outline_anz <- terra::crop(super_outline, anz_ext)
outline_sa <- terra::crop(super_outline, sa_ext)

if (!terra::same.crs(r_anz, outline_anz)) {
  outline_anz <- terra::project(outline_anz, r_anz)
}
if (!terra::same.crs(r_sa, outline_sa)) {
  outline_sa <- terra::project(outline_sa, r_sa)
}
```

```{r map-super, message=FALSE, fig.cap="Super Rugby grounds on cropped elevation: New Zealand + Australia (merge) and South Africa."}
p_anz <- ggplot() +
  tidyterra::geom_spatraster(data = r_anz, aes(fill = elevation)) +
  tidyterra::scale_fill_hypso_tint_c(
    name = "m",
    guide = guide_colorbar(barwidth = unit(2.6, "cm"), title.position = "top")
  ) +
  tidyterra::geom_spatvector(data = anz_pts, aes(colour = country), size = 2.8, show.legend = TRUE) +
  tidyterra::geom_spatvector_label(
    data = anz_pts,
    aes(label = club),
    size = 2,
    colour = "grey10",
    label.size = 0.12,
    show.legend = FALSE
  ) +
  tidyterra::geom_spatvector(data = outline_anz, fill = NA, colour = "grey25", linewidth = 0.25) +
  scale_colour_brewer(palette = "Set1", name = "Country") +
  labs(title = "Super Rugby — New Zealand & Australia") +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, margin = margin(b = 4)),
    legend.title = element_text(size = 8),
    legend.position = "bottom"
  )

p_sa <- ggplot() +
  tidyterra::geom_spatraster(data = r_sa, aes(fill = elevation)) +
  tidyterra::scale_fill_hypso_tint_c(
    name = "m",
    guide = guide_colorbar(barwidth = unit(2.6, "cm"), title.position = "top")
  ) +
  tidyterra::geom_spatvector(data = sa_pts, aes(colour = country), size = 2.8, show.legend = TRUE) +
  tidyterra::geom_spatvector_label(
    data = sa_pts,
    aes(label = club),
    size = 2,
    colour = "grey10",
    label.size = 0.12,
    show.legend = FALSE
  ) +
  tidyterra::geom_spatvector(data = outline_sa, fill = NA, colour = "grey25", linewidth = 0.25) +
  scale_colour_brewer(palette = "Set1", name = "Country") +
  labs(title = "Super Rugby — South Africa") +
  theme_void() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, margin = margin(b = 4)),
    legend.title = element_text(size = 8),
    legend.position = "bottom"
  )

p_anz + p_sa + patchwork::plot_layout(ncol = 2, widths = c(1, 1))
```

### Interactive maps

The maps below use **Leaflet** (via **`htmlwidgets`**). You can **pan**, **zoom**, and click stadium markers for **club**, **venue**, and **extracted elevation**. The elevation layer is a **downsampled** version of the same cropped DEM used in the static figures, so it loads quickly in the browser. An internet connection is required for the light **basemap** tiles.

```{r map-interactive-prep, echo=FALSE, message=FALSE}
flip_raster_east_west <- function(r) {
  mat <- terra::values(r, mat = TRUE)
  mat <- mat[, rev(seq_len(ncol(mat))), drop = FALSE]
  terra::values(r) <- mat
  r
}

leaflet_elev_map <- function(r, clubs_df) {
  r_small <- terra::aggregate(r, fact = 4, fun = mean)
  if (!grepl("4326", terra::crs(r_small, describe = TRUE)$code)) {
    r_small <- terra::project(r_small, "EPSG:4326")
  }
  r_small <- flip_raster_east_west(r_small)
  vals <- terra::values(r_small)
  vals <- vals[!is.na(vals)]
  elev_domain <- range(
    c(vals, clubs_df$ground_elevation_m),
    na.rm = TRUE
  )

  pal <- leaflet::colorNumeric(
    grDevices::terrain.colors(60),
    domain = elev_domain,
    na.color = "#808080"
  )

  # Plain scalars only: named bbox values break htmlwidgets JSON (fitBounds gets objects).
  lng1 <- terra::xmin(r_small)
  lng2 <- terra::xmax(r_small)
  lat1 <- terra::ymin(r_small)
  lat2 <- terra::ymax(r_small)
  bounds_ok <- all(is.finite(c(lng1, lng2, lat1, lat2))) &&
    lng1 < lng2 &&
    lat1 < lat2
  center_lng <- mean(clubs_df$lon, na.rm = TRUE)
  center_lat <- mean(clubs_df$lat, na.rm = TRUE)
  marker_fill <- unname(pal(clubs_df$ground_elevation_m))

  map <- leaflet::leaflet(
    width = "100%",
    height = 480,
    options = leaflet::leafletOptions(minZoom = 4)
  )

  map <- if (bounds_ok) {
    leaflet::fitBounds(map, lng1, lat1, lng2, lat2)
  } else {
    leaflet::setView(map, center_lng, center_lat, zoom = 4)
  }

  map |>
    leaflet::addProviderTiles(
      "CartoDB.Positron",
      group = "Basemap"
    ) |>
    leaflet::addRasterImage(
      r_small,
      colors = pal,
      opacity = 0.78,
      project = FALSE,
      group = "Elevation (SRTM)"
    ) |>
    leaflet::addLegend(
      pal = pal,
      values = vals,
      title = "Elevation (m)",
      position = "bottomright",
      opacity = 0.9
    ) |>
    leaflet::addCircleMarkers(
      lng = clubs_df$lon,
      lat = clubs_df$lat,
      radius = 8,
      stroke = TRUE,
      weight = 1.5,
      color = "#333333",
      fillColor = marker_fill,
      fillOpacity = 0.92,
      popup = paste0(
        "<strong>", clubs_df$club, "</strong><br>",
        clubs_df$stadium, "<br>",
        "<em>", clubs_df$country, "</em><br>",
        round(clubs_df$ground_elevation_m, 1), " m a.s.l."
      ),
      group = "Stadiums"
    ) |>
    leaflet::addLayersControl(
      baseGroups = "Basemap",
      overlayGroups = c("Elevation (SRTM)", "Stadiums"),
      options = leaflet::layersControlOptions(collapsed = FALSE)
    )
}

find_book_root <- function() {
  wd <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  for (steps in 0:6) {
    root <- if (steps == 0) {
      wd
    } else {
      normalizePath(
        file.path(wd, paste(rep("..", steps), collapse = "/")),
        winslash = "/",
        mustWork = FALSE
      )
    }
    if (file.exists(file.path(root, "_config.yml"))) {
      return(root)
    }
  }
  stop("Could not find Jupyter Book root (_config.yml).")
}

# Full HTML documents inside .output.text_html are invalid and browsers skip the maps.
# Save self-contained widgets under book _static/ and embed with iframes.
include_leaflet_map <- function(map, title, file_stem) {
  maps_dir <- file.path(find_book_root(), "_static", "uk-terrain-rugby", "maps")
  dir.create(maps_dir, recursive = TRUE, showWarnings = FALSE)
  out_path <- file.path(maps_dir, paste0(file_stem, ".html"))
  suppressWarnings(htmlwidgets::saveWidget(map, out_path, selfcontained = TRUE))
  iframe_src <- paste0("../../_static/uk-terrain-rugby/maps/", file_stem, ".html")
  iframe_html <- sprintf(
    paste0(
      "<h4 style=\"text-align:center;font-weight:bold;margin:0.6em 0 0.2em 0;\">%s</h4>",
      "<iframe src=\"%s\" title=\"%s\" width=\"100%%\" height=\"500\" ",
      "style=\"border:none;\"></iframe>"
    ),
    htmltools::htmlEscape(title),
    iframe_src,
    htmltools::htmlEscape(title)
  )
  IRdisplay::display_html(iframe_html)
  invisible(NULL)
}

anz_elev <- super_elev |> dplyr::filter(elev_source %in% c("NZL", "AUS"))
sa_elev <- super_elev |> dplyr::filter(elev_source == "ZAF")
```

```{r map-interactive, echo=FALSE, message=FALSE, warning=FALSE}
include_leaflet_map(
  leaflet_elev_map(r_anz, anz_elev),
  "Super Rugby — New Zealand & Australia (interactive)",
  "super_rugby_anz"
)
include_leaflet_map(
  leaflet_elev_map(r_sa, sa_elev),
  "Super Rugby — South Africa (interactive)",
  "super_rugby_sa"
)
```

Across **New Zealand and Australia**, most franchises sit on **low coastal or basin** terrain in the DEM (e.g. **Blues**, **Waratahs**, **Reds**), while **South Africa** again mixes **sea-level** venues with **Highveld** outliers—useful context when comparing the full competition to **Super Rugby Pacific** today, which no longer includes the South African sides.
